---
title: ""
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## Areal (lattice) data {#sec-areal_data}

In this practical we will:

-   Explore tools for areal spatial data wrangling and visualization.

-   Learn how to fit an areal model in `inlabru`

In areal data our measurements are summarised across a set of discrete, non-overlapping spatial units such as postcode areas, health board or pixels on a satellite image. In consequence, the spatial domain is a countable collection of (regular or irregular) areal units at which variables are observed. Many public health studies use data aggregated over groups rather than data on individuals - often this is for privacy reasons, but it may also be for convenience.

In the next example we are going to explore data on respiratory hospitalisations for Greater Glasgow and Clyde between 2007 and 2011. The data are available from the `CARBayesdata` R Package:

```{r}
#| message: false
#| warning: false
library(CARBayesdata)

data(pollutionhealthdata)
data(GGHB.IZ)
```

The `pollutionhealthdata` contains the spatiotemporal data on respiratory hospitalisations, air pollution concentrations and socio-economic deprivation covariates for the 271 Intermediate Zones (IZ) that make up the Greater Glasgow and Clyde health board in Scotland. Data are provided by the [Scottish Government](http://statistics.gov.scot.) and the available variables are:

-   `IZ`: unique identifier for each IZ.
-   `year`: the year were the measruments were taken
-   `observed`: observed numbers of hospitalisations due to respiratory disease.
-   `expected`: expected numbers of hospitalisations due to respiratory disease computed using indirect standardisation from Scotland-wide respiratory hospitalisation rates.
-   `pm10`: Average particulate matter (less than 10 microns) concentrations.
-   `jsa`: The percentage of working age people who are in receipt of Job Seekers Allowance
-   `price`: Average property price (divided by 100,000).

The `GGHB.IZ` data is a Simple Features (`sf`) object containing the spatial polygon information for the set of 271 Intermediate Zones (IZ), that make up of the Greater Glasgow and Clyde health board in Scotland ( @fig-GGC ).

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 4
#| fig-height: 4
#| fig-align: center
#| label: fig-GGC
#| fig-cap: "Greater Glasgow and Clyde health board represented by 271 Intermediate Zones"
#| purl: false

library(mapview)
mapview(GGHB.IZ)

```

### Maipulating and visualizing areal data

Let's start by loading useful libraries:

```{r}
#| message: false
#| warning: false
library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)   
library(mapview)
library(sf)

# load some libraries to generate nice map plots
library(scico)
```

The `sf` package allows us to work with vector data which is used to represent points, lines, and polygons. It can also be used to read vector data stored as a shapefiles.

First, lets combine both data sets based on the Intermediate Zones (IZ) variable using the `merge` function from `base` R, and select only one year of data

```{r}
resp_cases <- merge(GGHB.IZ %>%
                      mutate(space = 1:dim(GGHB.IZ)[1]),
                             pollutionhealthdata, by = "IZ")%>%
  dplyr::filter(year == 2007) 
```

In epidemiology, disease risk is usually estimated using Standardized Mortality Ratios (SMR). The SMR for a given spatial areal unit $i$ is defined as the ratio between the observed ( $Y_i$ ) and expected ( $E_i$ ) number of cases:

$$
SMR_i = \dfrac{Y_i}{E_i}
$$

A value $SMR > 1$ indicates that there are more observed cases than expected which corresponds to a high risk area. On the other hand, if $SMR<1$ then there are fewer observed cases than expected, suggesting a low risk area.

We can manipulate `sf` objects the same way we manipulate standard data frame objects via the `dplyr` package. Lets use the pipeline command `%>%` and the `mutate` function to calculate the yearly SMR values for each IZ:

```{r}
#| message: false
#| warning: false
resp_cases <- resp_cases %>% 
  mutate(SMR = observed/expected )
```

Now we use `ggplot` to visualize our data by adding a `geom_sf` layer and coloring it according to our variable of interest (i.e., SMR).

```{r}
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| code-fold: show
ggplot()+
  geom_sf(data=resp_cases,aes(fill=SMR))+
  scale_fill_scico(palette = "roma")
```

As with the other types of spatial modelling, our goal is to observe and explain spatial variation in our data. Generally, we are aiming to produce a smoothed map which summarises the spatial patterns we observe in our data.

### Spatial neighbourhood structures

A key aspect of any spatial analysis is that observations closer together in space are likely to have more in common than those further apart. This can lead us towards approaches similar to those used in time series, where we consider the spatial *closeness* of our regions in terms of a *neighbourhood structure*.

The function [`poly2nb()`](https://r-spatial.github.io/spdep/reference/poly2nb.html) of the `spdep` package can be used to construct a list of neighbors based on areas with contiguous boundaries (e.g., using Queen contiguity).

```{r}
#| message: false
#| warning: false

library(spdep)

W.nb <- poly2nb(GGHB.IZ,queen = TRUE)
W.nb
```

```{r}
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| purl: false
plot(st_geometry(GGHB.IZ), border = "lightgray")
plot.nb(W.nb, st_geometry(GGHB.IZ), add = TRUE)

```

::: callout-note
You could use the `snap` argument within `poly2nb` to set a distance at which the different regions centroids are consider neighbours.
:::

With this neighborhood matrix, we can then fit a conditional autoregressive (CAR) model. One of the most popular CAR approaches to model spatial correlation is the Besag model a.k.a. Intrinsic Conditional Autoregressive (ICAR) model.

The conditional distribution for $u_i$ given $\mathbf{u}_{-i} = (u_i,\ldots,u_{i-1},u_{i+1},\ldots,u_n)^T$ is

$$
u_i|\mathbf{u}_{-i} \sim N\left(\frac{1}{d_i}\sum_{j\sim i}u_j,\frac{1}{d_i\tau_u}\right)
$$

where $\tau_u$ is the precision parameter, $j\sim i$ denotes that $i$ and $j$ are neighbors, and $d_i$ is the number of neighbors. Thus, the mean of $u_i$ is equivalent to the the mean of the effects over all neighbours, and the precision is proportional to the number of neighbors. The joint distribution is given by:

$$
\mathbf{u}|\tau_u \sim N\left(0,\frac{1}{\tau_u}Q^{-1}\right),
$$

Where $Q$ denotes the structure matrix defined as

$$
Q_{i,j} = \begin{cases}
d_i, & i = j \\
-1, & i \sim j \\
0, &\text{otherwise}
\end{cases}
$$

This structure matrix directly defines the neighbourhood structure and is sparse. We can compute the adjacency matrix using the function `nb2mat()` in the `spdep` library. Then convert the adjacency matrix into the precision matrix $\mathbf{Q}$ of the CAR model as follows:

```{r}
library(spdep)
R <- nb2mat(W.nb, style = "B", zero.policy = TRUE)
diag = apply(R,1,sum)
Q = -R
diag(Q) = diag
```

The ICAR model accounts only for spatially structured variability and does not include a limiting case where no spatial structure is present. Therefore, it is typically combined with an additional unstructured random effect $z_i|\tau_z \sim N(0,\tau_{z}^{-1})$ . The resulting model $v_i = u_i + z_i$ is known as the Besag-York-Mollié model (BYM) which is an extension to the intrinsic CAR model that contains an i.i.d. model component.

### Fitting an ICAR model in `inlabru`

We fit a first model to the data where we consider a Poisson model for the observed cases.

**Stage 1** Model for the response $$
y_i|\eta_i\sim\text{Poisson}(E_i\lambda_i)
$$ where $E_i$ are the expected cases for area $i$.

**Stage 2** Latent field model $$
\eta_i = \text{log}(\lambda_i) = \beta_0 + u_i + z_i
$$ where

-   $\beta_0$ is a common intercept
-   $\mathbf{u} = (u_1, \dots, u_k)$ is a conditional Autoregressive model (CAR) with precision matrix $\tau_u\mathbf{Q}$
-   $\mathbf{z} = (z_1, \dots, z_k)$ is an unstructured random effect with precision $\tau_z$

**Stage 3** Hyperparameters

The hyperparameters of the model are $\tau_u$ and $\tau_z$

**NOTE** In this case the linear predictor $\eta$ consists of three components!!

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

Fit the above model in using `inlabru` by completing the following code:

```{r}
#| echo: true
#| eval: false


cmp = ~ Intercept(1) + space(...) + iid(...)

formula = ...


lik = bru_obs(formula = formula, 
              family = ...,
              E = ...,
              data = ...)

fit = bru(cmp, lik)

```

`r hide("Answer")`

```{r}
#| purl: false
cmp = ~ Intercept(1) + space(space, model = "besag", graph = Q) + iid(space, model = "iid")

formula = observed ~ Intercept + space + iid

lik = bru_obs(formula = formula, 
              family = "poisson",
              E = expected,
              data = resp_cases)

fit = bru(cmp, lik)

```

`r unhide()`
:::

After fitting the model we want to extract results.

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

1.  What is the estimated value for $\beta_0$?

2.  Look at the estimated values of the hyperparameters using

```{r}
#| echo: true
#| eval: false
fit$summary.hyperpar
```

Which of the two spatial components (structured or unstructured) explains more of the variability in the counts?
:::

### Areal model predictions

We now look at the predictions over space.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Complete the code below to produce:

1.  prediction of the linear predictor $\eta_i$

2.  The risk rate $\lambda_i$

3.  The expected cases $E_i\exp(\lambda_i)$ over the whole space of interest.

Plot the mean and sd of the resulting surfaces.

```{r}
#| echo: true
#| eval: false
#| purl: false

pred = predict(fit, resp_cases, ~data.frame(log_risk = ...,
                                             risk = exp(...),
                                             cases = ...
                                             ),
               n.samples = 1000)

```

`r hide("See Solution")`

```{r}
# produce predictions
pred = predict(fit, 
               resp_cases,
               ~data.frame(log_risk = Intercept + space,
                           risk = exp(Intercept + space),
                           cases = expected * exp(Intercept + space)),
               n.samples = 1000)

# plot the predictions

p1 = ggplot() + 
  geom_sf(data = pred$log_risk, aes(fill = mean)) + scale_fill_scico(direction = -1) +
  ggtitle("mean log risk")
p2 = ggplot() + 
  geom_sf(data = pred$log_risk, aes(fill = sd)) + scale_fill_scico(direction = -1) +
  ggtitle("sd log risk")

p3 = ggplot() +
  geom_sf(data = pred$risk, aes(fill = mean)) + scale_fill_scico(direction = -1) +
  ggtitle("mean  risk")

p4 = ggplot() + 
  geom_sf(data = pred$risk, aes(fill = sd)) +
  scale_fill_scico(direction = -1) +
  ggtitle("sd  risk")

p5 = ggplot() + geom_sf(data = pred$cases, aes(fill = mean)) + scale_fill_scico(direction = -1)+ 
  ggtitle("mean  expected counts")
p6 = ggplot() + geom_sf(data = pred$cases, aes(fill = sd)) + scale_fill_scico(direction = -1)+
  ggtitle("sd  expected counts")

p1 + p2 + p3 + p4 +p5 + p6 + plot_layout(ncol=2)

```

`r unhide()`
:::

Finally we want to compare our observations $y_i$ with the predicted means of the Poisson distribution $E_i\exp(\lambda_i)$

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-align: center
pred$cases %>% ggplot() + geom_point(aes(observed, mean)) + 
  geom_errorbar(aes(observed, ymin = q0.025, ymax = q0.975)) +
  geom_abline(intercept = 0, slope = 1)

```

Here we are predicting the *mean* of counts, not the counts. Predicting the counts is beyond the scope of this short course but you can check the supplementary material below.

::: {.callout-note appearance="simple" icon="false"}
## {{< bi book color=#298196 >}} Supplementary Material

Posterior predictive distributions, i.e., $\pi(y_i^{\text{new}}|\mathbf{y})$ are of interest in many applied problems. The `bru()` function does not return predictive densities. In the previous step we have computed predictions for the `expected counts` $\pi(E_i\lambda_i|\mathbf{y})$.

The predictive distribution is then: $$
\pi(y_i^{\text{new}}|\mathbf{y}) = \int \pi(y_i|E_i\lambda_i)\pi(E_i\lambda_i|\mathbf{y})\ dE_i\lambda_i
$$ where, in our case, $\pi(y_i|E_i\lambda_i)$ is Poisson with mean $E_i\lambda_i$. We can achieve this using the following algorithm:

1.  Simulate $n$ replicates of $g^k = E_i\lambda_i$ for $k = 1,\dots,n$ using the function `generate()` which takes the same input as `predict()`
2.  For each of the $k$ replicates simulate a new value $y_i^{new}$ using the function `rpois()`
3.  Summarise the $n$ samples of $y_i^{new}$ using, for example the mean and the 0.025 and 0.975 quantiles.
4.  

```{r}
#| fig-align: center
#| fig-width: 4
#| fig-height: 4
#| code-fold: true
#| code-summary: "Click here to see the code"

# simulate 1000 realizations of E_i\lambda_i
expected_counts = generate(fit, resp_cases, 
                           ~ expected * exp(Intercept + space),
                           n.samples = 1000)


# simulate poisson data
aa = rpois(271*1000, lambda = as.vector(expected_counts))
sim_counts = matrix(aa, 271, 1000)

# summarise the samples with posterior means and quantiles
pred_counts = data.frame(observed = resp_cases$observed,
                         m = apply(sim_counts,1,mean),
                         q1 = apply(sim_counts,1,quantile, 0.025),
                         q2 = apply(sim_counts,1,quantile, 0.975),
                         vv = apply(sim_counts,1,var)
                         )
# Plot the observations against the predicted new counts and the predicted expected counts

ggplot() + 
  geom_point(data = pred_counts, aes(observed, m, color = "Pred_obs")) + 
  geom_errorbar(data = pred_counts, aes(observed, ymin = q1, ymax = q2, color = "Pred_obs")) +
  geom_point(data = pred$cases, aes(observed, mean, color = "Pred_means")) + 
  geom_errorbar(data = pred$cases, aes(observed, ymin = q0.025, ymax = q0.975, color = "Pred_means")) +
  
  geom_abline(intercept = 0, slope =1)

```
:::
