---
title: "BES workshop"
author: "Jafet Belmont, Sara Martino and Janine Illian"
subtitle: "British Ecological Society Meeting 2025, Edinburgh"
format: 
  html: default
  PrettyPDF-pdf:
    keep-tex:  true
    number-sections: true
---

# Welcome to the course!

-   Welcome to the `inlabru` workshop!
-   The goal of this workshop is to introduce new users to the `inlabru` software for Bayesian spatial modelling using the integrated nested Laplace approximation (INLA).
-   The workshop is intended for applied statisticians and quantitative ecologists who are interested in efficiently modelling spatial data- No prior knowledge of `R-INLA` is required.

# Learning Objectives for the workshop

At the end of the workshop, participants will be able to:

-   Understand the `inlabru` workflow and software principles.
-   Fit commonly used models such as Linear models, LMM, GLM and GAMs using `inlabru`.
-   Describe different types of spatial data and fit appropriate spatial models.
-   Interpret model outputs and create spatial maps and predictions.
-   Implement a complete, reproducible spatial analysis workflow from data preparation to result visualization.

# Schedule Program

| Time | Topic |
|:----------------------------------:|:----------------------------------:|
| 13:00 - 14:00 | Workshop registration |
| 14:00 - 14:45 | [**Session 1**](https://ecol-stats.github.io/inlabru_workshop/BES_workshop/docs/slides/slides_1): Introduction to `inlabru` |
| 14:45 - 15:45 | [Practical Session 1](day1_practical.qmd) |
| 15:45 - 16:00 | Coffee break |
| 16:00 - 16:45 | [**Session 2**:](https://ecol-stats.github.io/inlabru_workshop/BES_workshop/docs/slides/slides_2) Spatial Modelling with `inlabru` |
| 16:45 - 17:45 | [Practical Session 2](practical2_compiler.qmd) |
| 17:45 - 18:00 | Wrap-up and outlook |

: {.striped .hover}

\clearpage

# In preparation for the workshop

Participants are required to follow the next steps before the day of the workshop:

1.  Install [R-INLA](https://www.r-inla.org/download-install)

2.  Install `inlabru` (available from CRAN)

    ```{r}
    #| eval: false

    # Enable universe(s) by inlabru-org
    options(repos = c(
      inlabruorg = "https://inlabru-org.r-universe.dev",
      INLA = "https://inla.r-inla-download.org/R/testing",
      CRAN = "https://cloud.r-project.org"
    ))

    # Install some packages
    install.packages("inlabru")
    ```

3.  Make sure you have the latest `R-INLA`, `inlabru` and `R` versions installed.

4.  Install the following libraries:

```{r}
#| echo: false
#| warning: false
#| message: false

    if (file.exists("practicals_libraries.rds")) {
      libs_data <- readRDS("practicals_libraries.rds")
      
      # Libraries to exclude
      exclude_libs <- c("webexercises", "INLA", "inlabru", "knitr")
      
      # Filter out excluded libraries
      filtered_libs <- libs_data[!libs_data$Library %in% exclude_libs, ]
      
      if (nrow(filtered_libs) > 0) {

        cat("install.packages(c(\n")
        cat(paste0('  "', filtered_libs$Library, '"', collapse = ",\n"))
        cat("\n))")
        
      } else {
        cat("No libraries found after filtering.\n")
      }
    }
```

