---
title: "Practical"
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## Linear Model {#sec-linmodel}

In this practical we will:

-   Simulate Gaussian data
-   Learn how to fit a linear model with `inlabru`
-   Generate predictions from the model

Start by loading useful libraries:

```{r}
#| warning: false
#| message: false
#| code-summary: "Load libraries"

library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
# load some libraries to generate nice plots
library(scico)
```

As our first example we consider a simple linear regression model with Gaussian observations $$
y_i\sim\mathcal{N}(\mu_i, \sigma^2), \qquad i = 1,\dots,N
$$

where $\sigma^2$ is the observation error, and the mean parameter $\mu_i$ is linked to the **linear predictor** ($\eta_i$) through an identity function: $$
\eta_i = \mu_i = \beta_0 + \beta_1 x_i
$$ where $x_i$ is a covariate and $\beta_0, \beta_1$ are parameters to be estimated. We assign $\beta_0$ and $\beta_1$ a vague Gaussian prior.

To finalize the Bayesian model we assign a $\text{Gamma}(a,b)$ prior to the precision parameter $\tau = 1/\sigma^2$ and two independent Gaussian priors with mean $0$ and precision $\tau_{\beta}$ to the regression parameters $\beta_0$ and $\beta_1$ (we will use the default prior settings in INLA for now).

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

What is the dimension of the hyperparameter vector and latent Gaussian field?

`r hide("Answer")`

The hyperparameter vector has dimension 1, $\pmb{\theta} = (\tau)$ while the latent Gaussian field $\pmb{u} = (\beta_0, \beta_1)$ has dimension 2, $0$ mean, and sparse precision matrix:

$$
\pmb{Q} = \begin{bmatrix}
\tau_{\beta_0} & 0\\
0 & \tau_{\beta_1}
\end{bmatrix}
$$ Note that, since $\beta_0$ and $\beta_1$ are fixed effects, the precision parameters $\tau_{\beta_0}$ and $\tau_{\beta_1}$ are fixed.

`r unhide()`
:::

::: callout-note
We can write the linear predictor vector $\pmb{\eta} = (\eta_1,\dots,\eta_N)$ as

$$
\pmb{\eta} = \pmb{A}\pmb{u} = \pmb{A}_1\pmb{u}_1 + \pmb{A}_2\pmb{u}_2 = \begin{bmatrix}
1 \\
1\\
\vdots\\
1
\end{bmatrix} \beta_0 + \begin{bmatrix}
x_1 \\
x_2\\
\vdots\\
x_N
\end{bmatrix} \beta_1
$$

Our linear predictor consists then of two components: an intercept and a slope.
:::

### Simulate example data

First, we simulate data from the model

$$
y_i\sim\mathcal{N}(\eta_i,0.1^2), \ i = 1,\dots,100
$$

with

$$
\eta_i = \beta_0 + \beta_1 x_i
$$

where $\beta_0 = 2$, $\beta_1 = 0.5$ and the values of the covariate $x$ are generated from an Uniform(0,1) distribution. The simulated response and covariate data are then saved in a `data.frame` object.

```{r}
#| code-fold: show
#| code-summary: "Simulate Data from a LM"

beta = c(2,0.5)
sd_error = 0.1

n = 100
x = rnorm(n)
y = beta[1] + beta[2] * x + rnorm(n, sd = sd_error)

df = data.frame(y = y, x = x)  

```

### Fitting a linear regression model with `inlabru`

------------------------------------------------------------------------

**Defining model components**

The model has two parameters to be estimated $\beta_1$ and $\beta_2$. We need to define the two corresponding model components.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Define an object called `cmp` that includes and (i) intercept `beta_0` and (ii) a covariate `x` linear effect `beta_1`.

`r hide("Take hint")`

The `cmp` object is here used to define model components. We can give them any useful names we like, in this case, `beta_0` and `beta_1`. You can remove the automatic intercept construction by adding a `-1` in the components

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
cmp =  ~ -1 + beta_0(1) + beta_1(x, model = "linear")

```
:::

::: callout-note
Note that we have excluded the default Intercept term in the model by typing `-1` in the model components. However, `inlabru` has automatic intercept that can be called by typing `Intercept()` , which is one of `inlabru` special names and it is used to define a global intercept, e.g.

```{r}
#| eval: false
#| purl: false

cmp =  ~  Intercept(1) + beta_1(x, model = "linear")

```
:::

**Observation model construction**

The next step is to construct the observation model by defining the model likelihood. The most important inputs here are the `formula`, the `family` and the `data`.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Define a linear predictor `eta` using the component labels you have defined on the previous task.

`r hide("Take hint")`

The `eta` object defines how the components should be combined in order to define the model predictor.

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
eta = y ~ beta_0 + beta_1

```
:::

The likelihood for the observational model is defined using the `bru_obs()` function.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Define the observational model likelihood in an object called `lik` using the `bru_obs()` function.

`r hide("Take hint")`

The `bru_obs` is expecting three arguments:

-   The linear predictor `eta` we defined in the previous task
-   The data likelihood (this can be specified by setting `family = "gaussian"`)
-   The data set `df`

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
lik =  bru_obs(formula = eta,
            family = "gaussian",
            data = df)

```
:::

**Fit the model**

We fit the model using the `bru()` functions which takes as input the components and the observation model:

```{r}
#| code-summary: "Fit LM in `inlabru`"
fit.lm = bru(cmp, lik)
```

**Extract results**

The `summary()` function will give access to some basic information about model fit and estimates

```{r }
#| code-summary: "Model summaries"
#| collapse: true
summary(fit.lm)
```

We can see that both the intercept and slope and the error precision are correctly estimated.

### Generate model predictions

------------------------------------------------------------------------

Now we can take the fitted `bru` object and use the `predict` function to produce predictions for $\mu$ given a new set of values for the model covariates or the original values used for the model fit

```{r}
new_data = data.frame(x = c(df$x, runif(10)),
                      y = c(df$y, rep(NA,10)))
pred = predict(fit.lm, new_data, ~ beta_0 + beta_1,
               n.samples = 1000)



```

The `predict` function generate samples from the fitted model. In this case we set the number of samples to 1000.

::: panel-tabset
## Plot

```{r}
#| code-fold: true
#| fig-cap: Data and 95% credible intervals
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

pred %>% ggplot() + 
  geom_point(aes(x,y), alpha = 0.3) +
  geom_line(aes(x,mean)) +
  geom_line(aes(x, q0.025), linetype = "dashed")+
  geom_line(aes(x, q0.975), linetype = "dashed")+
  xlab("Covariate") + ylab("Observations")
```

## R Code

```{r}
#| code-fold: show
#| eval: false
#| purl: false


pred %>% ggplot() + 
  geom_point(aes(x,y), alpha = 0.3) +
  geom_line(aes(x,mean)) +
  geom_line(aes(x, q0.025), linetype = "dashed")+
  geom_line(aes(x, q0.975), linetype = "dashed")+
  xlab("Covariate") + ylab("Observations")
```
:::

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Generate predictions for a new observation with $x_0 = 0.45$

`r hide("Take hint")`

You can create a new data frame containing the new observation $x_0$ and then use the `predict` function.

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

new_data = data.frame(x = 0.45)
pred = predict(fit.lm, new_data, ~ beta_0 + beta_1,
               n.samples = 1000)

```
:::
