---
title: "Practical 2"
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(webexercises)

```

## Setting priors and model checking for Linear Models

In this exercise we will:

-   Learn how to set priors for linear effects $\beta_0$ and $\beta_1$
-   Learn how to set the priors for the hyperparameter $\tau = 1/\sigma^2$.
-   Visualize marginal posterior distributions

Start by loading useful libraries:

```{r}
#| warning: false
#| message: false


library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
```

Recall a simple linear regression model with Gaussian observations $$
y_i\sim\mathcal{N}(\mu_i, \sigma^2), \qquad i = 1,\dots,N
$$

where $\sigma^2$ is the observation error, and the mean parameter $\mu_i$ is linked to the linear predictor through an identity function: $$
\eta_i = \mu_i = \beta_0 + \beta_1 x_i
$$ where $x_i$ is a covariate and $\beta_0, \beta_1$ are parameters to be estimated. In INLA, we assume that the model is a latent Gaussian model, i.e., we have to assign $\beta_0$ and $\beta_1$ a Gaussian prior. For the precision hyperparameter $\tau = 1/\sigma^2$ a typical prior choice is a $\text{Gamma}(a,b)$ prior.

In `R-INLA`, the default choice of priors for each $\beta$ is

$$
\beta \sim \mathcal{N}(0,10^3).
$$

and the prior for the variance parameter in terms of the log precision is

$$ \log(\tau) \sim \mathrm{logGamma}(1,5 \times 10^{-5}) $$

::: callout-note
If your model uses the default intercept construction (i.e., `Intercept(1)` in the linear predictor) `inlabru` will assign a default $\mathcal{N} (0,0)$ prior to it.
:::

Lets see how can we change the default priors using some simulated data

#### **Simulate example data**

We simulate data from a simple linear regression model

```{r}
#| code-fold: show
beta = c(2,0.5)
sd_error = 0.1

n = 100
x = rnorm(n)
y = beta[1] + beta[2] * x + rnorm(n, sd = sd_error)

df = data.frame(y = y, x = x)  

```

#### **Fitting the linear regression model with `inlabru`**

Now we fit a simple linear regression model in `inlabru` by defining (1) the model components, (2) the linear predictor and (3) the likelihood.

```{r }
# Model components
cmp =  ~ -1 + beta_0(1) + beta_1(x, model = "linear")
# Linear predictor
formula = y ~ Intercept + beta_1
# Observational model likelihood
lik =  bru_obs(formula = y ~.,
            family = "gaussian",
            data = df)
# Fit the Model
fit.lm = bru(cmp, lik)
```

### Change the prior distributions

Until now, we have used the default priors for both the precision $\tau$ and the fixed effects $\beta_0$ and $\beta_1$. Let's see how to customize these.

To check which priors are used in a fitted model one can use the function `inla.prior.used()`

```{r}
#| eval: true
#| purl: true 

inla.priors.used(fit.lm)
```

From the output we see that the precision for the observation $\tau\sim\text{Gamma}(1e+00,5e-05)$ while $\beta_0$ and $\beta_1$ have precision 0.001, that is variance $1/0.001$.

**Change the precision for the linear effects**

The precision for linear effects is set in the component definition. For example, if we want to increase the precision to 0.01 for $\beta_0$ we define the relative components as:

```{r }
#| eval: false 
#| purl: false

cmp1 =  ~-1 +  beta_0(1, prec.linear = 0.01) + beta_1(x, model = "linear")
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Run the model again using 0.1 as default precision for both the intercept and the slope parameter.

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: false
#| fig-width: 6 
#| fig-height: 4
#| fig-align: center 
#| purl: false
#| eval: true  
 
cmp2 =  ~ -1 + 
          beta_0(1, prec.linear = 0.1) + 
          beta_1(x, model = "linear", prec.linear = 0.1)

lm.fit2 = bru(cmp2, lik) 
```

Note that we can use the same observation model as before since both the formula and the dataset are unchanged.
:::

**Change the prior for the precision of the observation error** $\tau$

Priors on the hyperparameters of the observation model must be passed by defining argument `hyper` within `control.family` in the call to the `bru_obs()` function.

```{r}

# First we define the logGamma (0.01,0.01) prior 

prec.tau <- list(prec = list(prior = "loggamma",   # prior name
                             param = c(0.01, 0.01))) # prior values

lik2 =  bru_obs(formula = y ~.,
                family = "gaussian",
                data = df,
                control.family = list(hyper = prec.tau))

fit.lm2 = bru(cmp2, lik2) 

```

The names of the priors available in **R-INLA** can be seen with `names(inla.models()$prior)`

### Visualizing the posterior marginals

Posterior marginal distributions of the ﬁxed effects parameters and the hyperparameters can be visualized using the `plot()` function by calling the name of the component. For example, if want to visualize the posterior density of the intercept $\beta_0$ we can type:

```{r}
#| fig-width: 4
#| fig-align: center
#| fig-height: 4
#| code-fold: show

plot(fit.lm, "beta_0")
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Plot the posterior marginals for $\beta_1$ and for the precision of the observation error $\pi(\tau|y)$

`r hide("Take hint")`

See the `summary()` output to check the names for the different model components.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
plot(fit.lm, "beta_1") +
plot(fit.lm, "Precision for the Gaussian observations")
```
:::
